Visual Diagnostics of a Model Explainer -- Tools for the Assessment of LIME Explanations

### 2020-08-28 paper.Rnw
changes based on discussion in meeting

% Set up for ASA data science journal
\articletype{\kge{Research Article}}
\received{\kge{XX August 2020}}
\revised{\kge{XX XX XXXX}}
\accepted{\kge{XX XX XXXX}}
\raggedbottom

@@

% Abstract and keywords
\abstract[Summary]{The importance of providing explanations for predictions made by black-box models has led to the development of model explainer methods such as LIME (local interpretable model-agnostic explanations) \citep{ribeiro:2016}. LIME uses a surrogate model to explain the relationship between predictor variables and predictions from a black-box model in a local region around a prediction of interest. However, the quality of the resulting explanations relies on how well the explainer model captures the black-box model in a specified local region. Here we introduce three visual diagnostics to assess the quality of LIME explanations: (1) explanation scatterplots, (2) assessment metric plots, and (3) feature heatmaps. We apply the visual diagnostics to a forensics bullet matching dataset to show examples where LIME explanations depend on the tuning parameters and the explainer model oversimplifies the black-box model. Our examples raise concerns about claims made of LIME \citep{ribeiro:2016} that are similar to other criticisms in the literature \citep{alvarezmelis:2018, laugel:2018, molnar:2019}.}
\keywords{LIME, black-box models, interpretability, diagnostics}

@@

The concept of LIME is relatively simple: use an interpretable model to approximate a complex model in a local region. However, a practical implementation of LIME is not straightforward, and research is being done to improve the procedure \citep{laugel:2018}. The current implementations of LIME \citep{ribeiro:2020} \kgc{Is there a better way to cite a Python package?} \hh{might be good to model the python package reference after the R package references, have a loog at reference limepy} \kgc{I kept thinking about this, and since the Python pakcgae doesn't have a manual, I ended up combining your limepy reference with the second second suggestion from https://academia.stackexchange.com/questions/14010/how-do-you-cite-a-github-repository. For now, I listed Ribeiro as the author, because he is the main contributor, but there are 40 other contributors to the package. Is it okay to just list Ribeiro?}\hh{I would go with 'and contributors' - if we just cite Ribeiro people might not realize that it's truly not just him.} \citep{pedersen:2020} offer various tuning parameters (see Section \ref{background}) that affect the explainer model and ultimately, the explanation. Since the explainer model is an approximation of the complex model and not a direct interpretation, the explanations produced by an explainer model are subject to the quality of the approximation. Thus, in order to achieve accurate explanations, the tuning parameters selected need to be assessed.

@@

\citet{ribeiro:2016} provide an implementation of LIME in a Python package \cite{limepy}. An adaption of the Python package in R has been implemented and made available by Thomas Lin-Pedersen \citep{pedersen:2020}. The discussion of parameter choices and implementation details of LIME in this paper are based on the R package.

@@

\begin{enumerate}
\item \emph{Explanation Scatterplot} (Section \ref{exp-scatter}): Comparison of the explainer and complex models for an individual prediction of interest.
\item \emph{Feature Heatmap} (Section \ref{feat-heat}): Comparison of features selected by LIME across applications of LIME with different tuning parameters.
\item \emph{Assessment Metric Plot} (Section \ref{assess-metric}): Comparison of performance metrics for LIME across applications of LIME with different tuning parameters. 
\end{enumerate}

@@

We apply LIME using six sets of tuning parameters to all observations in the \data \ test set to observe variability across tuning parameters. A quantile bin based simulation method (samples are simulated uniformly from a specified number of quantile bins) is used for five of the LIME applications with the number of bins varying from 2 to 6 by application. We use 6 bins as the maximum, because the complexity of the explanations increases with the number of bins. Note that 4 quantile bins are the default method in \emph{lime}. The sixth application of LIME uses a kernel density simulation method (samples are drawn from kernel density approximations of the feature distributions). The default methods for feature selection (forward selection) and the computation of the weights (Gower distance raised to an exponent of 1) are used for all applications. (See Appendix~\ref{lime-details} for descriptions of the tuning parameters.)

@@

plot_grid(sine_poi_exp, exp_scatter) 
@
\caption{(left) Visualization from the \emph{lime} R package of a LIME explanation for the \data \ prediction of interest identified in \autoref{fig:sine-plot-data}. (right) \emph{Explanation Scatterplot}. The points represent the simulated data colored by the random forest probabilities and sized by the assigned LIME weights. %\hh{XXX could you make the complex prediction **much** lighter for this example? The diamond and the lines are hard to see, but should be the focus.} \kgc{Is that light enough? Feel free to edit the alpha value in the plot\_explain\_scatter function as you feel necessary. I also made the lines depicting the explainer model wider in limeaid, and there is an option 'line\_size' that can be adjusted.} \hh{thank you!}
The prediction of interest is shown as the diamond shaped point. The explainer model indicator variables associated with $x_1$ and $x_2$ (quantile bins containing the prediction of interest) are depicted by the solid lines. The line colors represent the explainer model coefficient signs and indicate whether the feature supports a prediction of `blue' (blue lines) or supports a prediction of `red' (red lines). This figure shows that the quantile bins are not flexible enough to capture the relationship between the random forest predictions and the $x_1$ and $x_2$ values.}
\label{fig:sine-exp-scatter}
\end{figure*}

%\subsection{A LIME explanation}

@@

%\kgc{I really like the idea of separating the description of the \emph{lime} figure from the explanation scatterplot, but the more I think about it, I don't like that it get it's own sub-section. I think it would be nicer to allocate the sub-sections to the visual diagnostic plots. Instead, I switched to using a heading like you created for introducing the sine data. I also changed the wording of the heading a bit. What do you think? Another option would be to drop the heading altogether and flow into the next paragraph.} \hh{XXX I've just pulled the description of the left hand side of figure 4 out of the next section.}
\hh{XXX I like the idea of not giving this section an official header, but I like the way it looks like now with the bold header - that should help the reader what's going on.}

Before introducing the visual diagnostics, let us consider a commonly used visualization of a LIME explanation that is shown in Figure \ref{fig:sine-exp-scatter} (left). The plot is created using the \emph{lime} R package and provides a visual representation of a LIME explanation for the prediction of interest indicated in \autoref{fig:sine-plot-data}. In particular, the explanation is obtained using 4 quantile bins to simulate the data. %\kgc{I moved the next few sentences here from the sub-section.} 
In this scenario, LIME converts continuous predictor variables to indicator variables identifying whether the variable value falls in the same quantile bin as the prediction of interest or not. The indicator variables are used as the explainer model features.

@@

The explanation for the prediction of interest depicted in Figure \ref{fig:sine-exp-scatter} is interpreted as follows. A random forest classification of blue is supported by the prediction of interest having a value of $x_2$ that is greater than \Sexpr{str_split(sine_poi_lime_default$feature_desc[[1]], pattern = " ")[[1]][1]}, but the prediction of interest having a value of $x_1$ that is greater than \Sexpr{str_split(sine_poi_lime_default$feature_desc[[2]], pattern = " ")[[1]][1]} and less than or equal to \Sexpr{str_split(sine_poi_lime_default$feature_desc[[2]], pattern = " ")[[1]][5]} provides support against a classification of blue. Since the weight associated with $x_2$ has a larger magnitude than the weight associated with $x_1$, LIME explains that $x_2$ plays a more important role in the random forest prediction. Additionally, consider that since the support for blue by $x_2$ outweights the support for red by $x_1$, the explainer model overall favors a label of blue for the prediction of interest, which agrees with the random forest probability of \Sexpr{sine_poi_lime_default$label_prob[1]} for blue. Note that both models classify the observation incorrectly.

%\hh{Because the support for blue by the value of feature $x_1$ outweighs the support for red by the value of feature $x_2$, the overall LIME explanation favors a label of blue for the prediction of interest.} \kgc{I don't think this statement is quite right.}

%Since the weight associated with $x_2$ has a larger magnitude than the weight associated with $x_1$, LIME explains that the $x_1$ and $x_2$ locations of the prediction of interest provide more support for a random forest classification of blue over red.

\subsection{Explanation Scatterplots} \label{exp-scatter}
%\hh{XXX I think we could reduce some of the example for the specific example a bit, now that the LIME explanation is moved out, there is not that much need to repeat the general ideas.} \kgc{I commented out the text that I think you were referring to. See what you think.}

%\kgc{I moved the next sentence here from the previous paragraph.} 
Based on the plot on the left of \autoref{fig:sine-exp-scatter} alone, it is not possible to make an informed assessment of the explanation. For a further assessment of the explainer model, we turn to an \emph{explanation scatterplot}:  a visual diagnostic for assessing the LIME claims of locality and fidelity for an individual explanation by juxtaposing the complex and explainer models in one plot. The format of an explanation scatterplot depends on the LIME simulation method. We introduce the explanation scatterplot here under the \emph{lime} R package default method of four quantile bins.

@@

An explanation scatterplot corresponding to the LIME explanation depicted on the left in \autoref{fig:sine-exp-scatter} is shown on the right hand side of \autoref{fig:sine-exp-scatter}.
%The points are the simulated values of $x_2$ versus $x_1$ colored by the random forest predictions and sized by the proximity weight used to fit the ridge regression explainer model. The explainer model is represented by the solid horizontal and vertical lines identifying the boundaries of the quantiles that contain the prediction of interest. The color of the lines indicate whether the explainer model coefficient is positive (blue) or negative (red). Thus, the $x_2$ coefficient is positive, so the $x_2$ value of the prediction of interest supports a random forest prediction of `blue', and the $x_1$ coefficient is negative, so the $x_1$ value of the prediction of interest contradicts a random forest prediction of `blue'. The prediction of interest is represented by the diamond shaped point.
By juxtaposing the random forest predictions and the explainer model boundaries, we are able to assess the faithfulness and localness of the explainer model.

@@

%\kgc{I tried rewriting this paragraph to make it less "technical". See what you think.}\hh{I like it - the arguments are also more to the point }
Now, consider the claim of faithfulness: It can be said that the majority of the points in the $x_2$ quantile support a prediction of blue, which is captured by the bar supporting a prediction of blue in the LIME explanation, and a similar statement can be made about the $x_1$ quantile bin. These statements validate the explanation produced by LIME. However, the explanation scatterplot plot shows that the random forest performs well at capturing the sine curve classification boundary by creating various sized rectangles consisting of predictions with similar probabilities, which the LIME explanation does not pick up on. Thus, LIME does provide an explanation for the prediction, but it is a very poor explanation in terms of faithfulness to the random forest prediction regions.

%\sout{Within the intersection of the $x_1$ and $x_2$ quantile bins, there are more random forest probabilities above 0.5 \kge{than below}. This indicates why the magnitude of the coefficient associated with $x_2$ is larger than that of $x_1$. However, the pattern of the random forest predictions in this plot suggest that better explanations for this prediction exist. One example of a better explanation is to say that the prediction of interest received a random forest prediction greater than 0.5 since the observation of interest falls in the region where $x_1$ is less than 2.5 and $x_2$ is greater 4.5. A more localized explanation is to say that the region around the prediction of interest shows that observations with $-0.3 < x_1 <= 2.5$ and $4.8 < x_2 <= 8.75$ are all assigned probabilities greater than 0.5. The procedure implemented by LIME is not flexible enough to capture either of these explanations.}

%\sout{The explanation scatterplot shows issues with a definition of locality and an oversimplification of the random forest \kge{\sout{model}}. LIME does provide an explanation for the prediction of interest, which makes sense based on the quantile bins used. However, the explanation scatterplot identifies that better explanations for the prediction of interest exist.}

@@

@
\caption{Hypothetical examples of feature heatmaps in two possible situations. The heatmaps show the top feature chosen for 10 cases across 5 different sets of  tuning parameters. The color of the cell indicates the feature chosen by LIME. Situation 1 is the ideal, because the explanations vary across cases but do not dependent on specific tuning parameter values.}
\label{fig:concept-feat-heat}

@@

<<sine-feat-heat, out.width = '0.45\\textwidth', fig.width = 4, fig.height = 4>>=
plot_feature_heatmap(sine_lime_explain$explain,
                order_method = "PCA") +
  theme_bw(base_family = "Times", base_size = 10) +
  theme(legend.position = "bottom",
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        strip.background = element_rect(color = "white", fill = "white")) +
  labs(y = "Case")
@
\caption{\emph{Feature Heatmap}. An example feature heatmap of the explanations from the applications of LIME with different tuning parameters to the \data \ test set. The cases from the test set are plotted on the y-axis, and the tuning parameter values (simulation methods here) are included on the x-axis. The colors of the tiles indicate the feature selected by LIME for the corresponding case and tuning parameter value. The plot is faceted by the first and second most important features selected by LIME (top and bottom facets, respectively). The vertical facets separate the  kernel density method from the quantile bin methods. The vertical striping indicates that the LIME explanations are not consistent across tuning parameters.}
\label{fig:sine-feat-heat}
\end{figure}

Explanations produced by LIME are likely to be affected by the choice of tuning parameters. A hypothetical example of this is shown by Figures \ref{fig:concept-plot-good} and \ref{fig:concept-plot-bad} where the method used to weight the observations influenced the explanation. As of the time of writing this manuscript, we have encountered no recommendations for how to specify the parameter values besides for the default settings in \emph{lime}. In order to compare the explanations produced by LIME using different tuning parameters and provide another perspective for assessing localness, we visualize an overview of the explanations with the \emph{feature heatmap} diagnostic plot.

The feature heatmap uses colors to identify the features selected by LIME across multiple predictions (referred to as cases here) and tuning paramter values organized by the  feature importance assigned by LIME. That is, for LIME applied with $t$ sets of tuning parameters to $n$ cases to select the $f$ top features, create $f$ heatmaps (one for each of the positions of importance determined by the magnitude of the explainer model coefficients) with the cases on the $y$-axis, the tuning parameters on the $x$-axis, and the cells colored by the feature chosen for the corresponding case and tuning parameter. Additional tuning parameters may also be included in the plot via facets.

Two hypothetical examples of feature heatmaps are included in \autoref{fig:concept-feat-heat}. The plots are created with the assumption that LIME is applied to select the top feature out of $p=\kge{4}$  features for $n=10$ cases with $t=5$ sets of tuning parameters. Situation 1 is  an example where the features selected are consistent across tuning parameters within a case but vary across cases within a tuning parameter. This is the ideal situation, because the LIME explanations do not depend on the tuning parameters but do depend on the location of the observation in the feature space. Situation 2 is an example where the selected features  vary across tuning parameters within a case but are consistent across cases within a tuning parameter. This situation indicates that the features selected by LIME are dependent on the tuning parameters, and the explanations may not be  local, because the same feature is chosen regardless of the case. In practice, it is expected that the plot will exhibit a combination of these two situations.

\autoref{fig:sine-feat-heat} shows a feature heatmap for the LIME applications to the 100 observations in the \data \ test set. The most important and second most important features selected by lime are showin in the top and bottom facets, respectively. For the quantile bins, the original features prior to the indicator variable transformation are included since it is obvious that different features would be selected when the sizes of the bins change. This figure shows that for kernel density and 2 quantile bins, LIME selects $x_1$ as the most important feature and $x_2$ as the second most important feature across all cases in the test set. These explanations are not local. There is variability in the features selected by LIME for 3 to 6 quantile bins suggesting more local explanations. There are signs of vertical striping, which suggests a dependence on tuning parameters. Note that the explanations from 3 and 5 quantile bins include the selection of the random noise variable ($x_3$) as an important variable in many predictions, which should not be the case. The pattern seen in the explanations for 3 to 6 quantile bins may suggest local explanations, but the dependence on tuning parameter makes it unclear which set of explanations to use.

@@

The feature heatmap for the \data \ in the previous section shows an example of inconsistent LIME explanations across tuning parameters. In this situation, the user must determine which set of explanations to trust. One way to do this is to compute assessment metrics for each set of explanations to identify the optimal tuning parameters. We discuss three metrics for this purpose and present a visual comparison in an \emph{assessment metric plot}.

Each metric presented below is computed on  LIME explanations for a set of predictions obtained using the same tuning parameters. Here we provide a high level description of the metrics. Notation and formulas for these metrics are included in Appendix \ref{metric-details}.

@@

\includegraphics[width=3.25in]{./figure-static/signatures.png}
\caption{Example bullet signatures. The bullet signatures correspond to the same land \hh{XXX barrel land is actually a standing term, meaning the physical land in the barrel that created the striation. Obviously it is confusing, so I got rid of the 'barrel'.} and therefore have very similar patterns. The \citet{hare:2016} random forest is fit using various features that measure the similarity between two such signatures.}
\label{fig:signatures}
\end{figure}

@@

We apply the random forest  \kge{\sout{model}} \hh{XXX I would actually prefer to leave the model in here. applying a random forest sounds like you applied he methodology to the data. But you got predictions based on the model previously fitted by Eric. That's a difference} from \citet{hare:2016} to another set of bullets from the Hamby study with \Sexpr{dim(bullet_test)[1]} rows of land comparisons. We first consider a global visualization of the relationship between the random forest predictions and the model features with  a parallel coordinate plot of the training data (top facet) and testing data (bottom facets) predictions (\autoref{fig:bullet-plot-rfpcp}). The majority of same source observations with random forest probabilities close to 1 have a clear pattern of corresponding feature values. The known same source observations where the random forest is wrong (does not return probabilities close to 1) have feature values that reflect those of observations where same source is known to be false.

LIME is applied to all test set observations using different tuning parameters: 12 sampling methods (2-6 equally spaced bins, 2-6 quantile bins, kernel density estimation, and normal approximation) and 3 Gower exponents (0.5, 1, and 10). Thus, a total of $12\times 3=36$ different applications of LIME are performed. We specify that each LIME explanation return 3 features and feature selection is performed by selecting the features with the highest weights in a ridge regression model (default option).

@@

First, the density simulation methods produce the same explanations for almost all cases and LIME tuning parameters suggesting the LIME explanations are global and not local. Second, within a bin based simulation method, the features selected by LIME for an observation often vary by the number of bins but do not appear to vary by the Gower power. With the equal bins, there are vertical stripes that suggest a dependence of the LIME explanations on the number of bins. The vertical stripes are not as apparent with the quantile bins. Lastly, there are clear differences between the LIME explanations produced by the bin based simulation methods for the matches and non-matches. This suggests that the features the random forest uses to classify a match or non-match are different.

@@

@
\caption{Feature heatmap of  36 LIME applications to the bullet comparison data test set. In addition to faceting the results by simulation method and LIME feature selection order, facets for the Gower power and whether the observation is a match or non-match are included. The vertical stripes of features selected indicate a dependence between the LIME explanations and tuning parameters.}
\label{fig:bullet-feature-heatmap}

@@

To try to identify a set of LIME tuning parameters with the most trustworthy set of explanations, an assessment metric plot is considered (\autoref{fig:bullet-assess-metric}). All density simulation methods performed well based on the three metrics. This is an interesting result, since \autoref{fig:bullet-feature-heatmap} shows that the density methods results in global explanations. The metric values for the bin based methods do not agree across metrics. For example, 2 and 3 quantile bins performed well according to the average $R^2$ but poorly according to MSEE. As a result, with the bin based methods, it is difficult to know which method to recommend. The applications using a power of 0.5 perform the best or as well as the other powers across all simulation methods suggesting that the power that leads to a more global explanation is preferred by LIME.

Perhaps we can say that the kernel density method, 3 equal bins, or 3 quantile bins are some of the better performing methods, but unfortunately, \autoref{fig:bullet-assess-metric} leaves us without a clear indication of which set of explanations to use to explain the random forest. To better understand the explanations provided by these three methods, we take a closer look at explanations from these methods for two observations of interest, which will be referred to as case M (a known match) and case NM (a known non-match).

>>>>

% Set up for ASA data science journal
\articletype{Research Article}
\received{XX XX 2020}
\revised{XX XX XXXX}
\accepted{XX XX XXXX}
\raggedbottom

@@

% Abstract and keywords
\abstract[Summary]{The importance of providing explanations for predictions made by black-box models has led to the development of model explainer methods such as LIME (local interpretable model-agnostic explanations) \citep{ribeiro:2016}. LIME uses a surrogate model to explain the relationship between predictor variables and predictions from a black-box model in a local region around a prediction of interest. However, the quality of the resulting explanations relies on how well the explainer model captures the black-box model in a specified local region. Here we introduce three visual diagnostics to assess the quality of LIME explanations: (1) explanation scatterplots, (2) assessment metric plots, and (3) feature heatmaps. We apply the visual diagnostics to a forensics bullet matching dataset to show examples where LIME explanations depend on the tuning parameter values and the explainer model oversimplifies the black-box model. Our examples raise concerns about claims made of LIME \citep{ribeiro:2016} that are similar to other criticisms in the literature \citep{alvarezmelis:2018, laugel:2018, molnar:2019}.}
\keywords{LIME, black-box models, interpretability, diagnostics}

@@

The concept of LIME is relatively simple: use an interpretable model to approximate a complex model in a local region. However, a practical implementation of LIME is not straightforward, and research is being done to improve the procedure \citep{laugel:2018}. The current implementations of LIME \citep{ribeiro:2020} \citep{pedersen:2020} offer various tuning parameters (see Section \ref{background}) that affect the explainer model and ultimately, the explanation. Since the explainer model is an approximation of the complex model and not a direct interpretation, the explanations produced by an explainer model are subject to the quality of the approximation. Thus, in order to achieve accurate explanations, the tuning parameter values selected need to be assessed.

@@

\citet{ribeiro:2016} provide an implementation of LIME in a Python package \cite{ribeiro:2020}. An adaption of the Python package in R has been implemented and made available by Thomas Lin-Pedersen \citep{pedersen:2020}. The discussion of parameter choices and implementation details of LIME in this paper are based on the R package.

@@

\begin{enumerate}
\item \emph{Explanation Scatterplot} (Section \ref{exp-scatter}): Comparison of the explainer and complex models for an individual prediction of interest.
\item \emph{Feature Heatmap} (Section \ref{feat-heat}): Comparison of features selected by LIME across applications of LIME with different tuning parameter values.
\item \emph{Assessment Metric Plot} (Section \ref{assess-metric}): Comparison of performance metrics for LIME across applications of LIME with different tuning parameter values. 
\end{enumerate}

@@

We apply LIME using six sets of tuning parameter values to all observations in the \data \ test set to observe variability across tuning parameter values. A quantile bin based simulation method (samples are simulated uniformly from a specified number of quantile bins) is used for five of the LIME applications with the number of bins varying from 2 to 6 by application. We use 6 bins as the maximum, because the complexity of the explanations increases with the number of bins. Note that 4 quantile bins are the default method in \emph{lime}. The sixth application of LIME uses a kernel density simulation method (samples are drawn from kernel density approximations of the feature distributions). The default methods for feature selection (forward selection) and the computation of the weights (Gower distance raised to an exponent of 1) are used for all applications. (See Appendix~\ref{lime-details} for descriptions of the tuning parameters.)

@@

plot_grid(sine_poi_exp, exp_scatter) 
@
\caption{(left) Visualization from the \emph{lime} R package of a LIME explanation for the \data \ prediction of interest identified in \autoref{fig:sine-plot-data}. (right) \emph{Explanation Scatterplot}. The points represent the simulated data colored by the random forest probabilities and sized by the assigned LIME weights. The prediction of interest is shown as the diamond shaped point. The explainer model indicator variables associated with $x_1$ and $x_2$ (quantile bins containing the prediction of interest) are depicted by the solid lines. The line colors represent the explainer model coefficient signs and indicate whether the feature supports a prediction of `blue' (blue lines) or supports a prediction of `red' (red lines). This figure shows that the quantile bins are not flexible enough to capture the relationship between the random forest predictions and the $x_1$ and $x_2$ values.}
\label{fig:sine-exp-scatter}
\end{figure*}

@@

Before introducing the visual diagnostics, let us consider a commonly used visualization of a LIME explanation that is shown in Figure \ref{fig:sine-exp-scatter} (left). The plot is created using the \emph{lime} R package and provides a visual representation of a LIME explanation for the prediction of interest indicated in \autoref{fig:sine-plot-data}. In particular, the explanation is obtained using 4 quantile bins to simulate the data. In this scenario, LIME converts continuous predictor variables to indicator variables identifying whether the variable value falls in the same quantile bin as the prediction of interest or not. The indicator variables are used as the explainer model features.

@@

The explanation for the prediction of interest depicted in Figure \ref{fig:sine-exp-scatter} is interpreted as follows. A random forest classification of blue is supported by the prediction of interest having a value of $x_2$ that is greater than \Sexpr{str_split(sine_poi_lime_default$feature_desc[[1]], pattern = " ")[[1]][1]}, but the prediction of interest having a value of $x_1$ that is greater than \Sexpr{str_split(sine_poi_lime_default$feature_desc[[2]], pattern = " ")[[1]][1]} and less than or equal to \Sexpr{str_split(sine_poi_lime_default$feature_desc[[2]], pattern = " ")[[1]][5]} provides support against a classification of blue. Since the weight associated with $x_2$ has a larger magnitude than the weight associated with $x_1$, LIME explains that $x_2$ plays a more important role in the random forest prediction. Additionally, consider that since the support for blue by $x_2$ outweighs the support for red by $x_1$, the explainer model overall favors a label of blue for the prediction of interest, which agrees with the random forest probability of \Sexpr{sine_poi_lime_default$label_prob[1]} for blue. Note that both models classify the observation incorrectly.

\subsection{Explanation Scatterplots} \label{exp-scatter}

Based on the plot on the left of \autoref{fig:sine-exp-scatter} alone, it is not possible to make an informed assessment of the explanation. For a further assessment of the explainer model, we turn to an \emph{explanation scatterplot}: a visual diagnostic for assessing the LIME claims of locality and fidelity for an individual explanation by juxtaposing the complex and explainer models in one plot. The format of an explanation scatterplot depends on the LIME simulation method. We introduce the explanation scatterplot here under the \emph{lime} R package default method of four quantile bins.

@@

An explanation scatterplot corresponding to the LIME explanation depicted on the left in \autoref{fig:sine-exp-scatter} is shown on the right hand side of \autoref{fig:sine-exp-scatter}. By juxtaposing the random forest predictions and the explainer model boundaries, we are able to assess the faithfulness and localness of the explainer model.

@@

Now, consider the claim of faithfulness: It can be said that the majority of the points in the $x_2$ quantile support a prediction of blue, which is captured by the bar supporting a prediction of blue in the LIME explanation, and a similar statement can be made about the $x_1$ quantile bin. These statements validate the explanation produced by LIME. However, the explanation scatterplot plot shows that the random forest performs well at capturing the sine curve classification boundary by creating various sized rectangles consisting of predictions with similar probabilities, which the LIME explanation does not pick up on. Thus, LIME does provide an explanation for the prediction, but it is a very poor explanation in terms of faithfulness to the random forest prediction regions.

@@

@
\caption{Hypothetical examples of feature heatmaps in two possible situations. The heatmaps show the top feature chosen for 10 cases across 5 different sets of tuning parameter values. The color of the cell indicates the feature chosen by LIME. Situation 1 is the ideal, because the explanations vary across cases but do not dependent on specific tuning parameter values.}
\label{fig:concept-feat-heat}

@@

<<sine-feat-heat, out.width = '0.45\\textwidth', fig.width = 4, fig.height = 4.5>>=
plot_feature_heatmap(sine_lime_explain$explain,
                order_method = "PCA") +
  theme_bw(base_family = "Times", base_size = 10) +
  theme(legend.position = "bottom",
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        strip.background = element_rect(color = "white", fill = "white")) +
  labs(y = "Case")
@
\caption{\emph{Feature Heatmap}. An example feature heatmap of the explanations from the applications of LIME with different tuning parameter values to the \data \ test set. The cases from the test set are plotted on the y-axis, and the tuning parameter values (simulation methods here) are included on the x-axis. The colors of the tiles indicate the feature selected by LIME for the corresponding case and tuning parameter value. The plot is faceted by the first and second most important features selected by LIME (top and bottom facets, respectively). The vertical facets separate the  kernel density method from the quantile bin methods. The vertical striping indicates that the LIME explanations are not consistent across tuning parameter values.}
\label{fig:sine-feat-heat}
\end{figure}

Explanations produced by LIME are likely to be affected by the choice of tuning parameter values. A hypothetical example of this is shown by Figures \ref{fig:concept-plot-good} and \ref{fig:concept-plot-bad} where the method used to weight the observations influenced the explanation. As of the time of writing this manuscript, we have encountered no recommendations for how to specify the parameter values besides for the default settings in \emph{lime}. In order to compare the explanations produced by LIME using different tuning parameter values and provide another perspective for assessing localness, we visualize an overview of the explanations with the \emph{feature heatmap} diagnostic plot.

The feature heatmap uses colors to identify the features selected by LIME across multiple predictions (referred to as cases here) and tuning parameter values organized by the  feature importance assigned by LIME. That is, for LIME applied with $t$ sets of tuning parameter values to $n$ cases to select the $f$ top features, create $f$ heatmaps (one for each of the positions of importance determined by the magnitude of the explainer model coefficients) with the cases on the $y$-axis, the tuning parameter values on the $x$-axis, and the cells colored by the feature chosen for the corresponding case and tuning parameter value. Additional tuning parameters may also be included in the plot via facets.

Two hypothetical examples of feature heatmaps are included in \autoref{fig:concept-feat-heat}. The plots are created with the assumption that LIME is applied to select the top feature out of $p=4$  features for $n=10$ cases with $t=5$ sets of tuning parameter values. Situation 1 is  an example where the features selected are consistent across tuning parameter values within a case but vary across cases within a tuning parameter value. This is the ideal situation, because the LIME explanations do not depend on the tuning parameters but do depend on the location of the observation in the feature space. Situation 2 is an example where the selected features vary across tuning parameter values within a case but are consistent across cases within a tuning parameter value. This situation indicates that the features selected by LIME are dependent on the tuning parameters, and the explanations may not be  local, because the same feature is chosen regardless of the case. In practice, it is expected that the plot will exhibit a combination of these two situations.

\autoref{fig:sine-feat-heat} shows a feature heatmap for the LIME applications to the 100 observations in the \data \ test set. The most important and second most important features selected by lime are shown in in the top and bottom facets, respectively. For the quantile bins, the original features prior to the indicator variable transformation are included since it is obvious that different features would be selected when the sizes of the bins change. This figure shows that for kernel density and 2 quantile bins, LIME selects $x_1$ as the most important feature and $x_2$ as the second most important feature across all cases in the test set. These explanations are not local. There is variability in the features selected by LIME for 3 to 6 quantile bins suggesting more local explanations. There are signs of vertical striping, which suggests a dependence on tuning parameters. Note that the explanations from 3 and 5 quantile bins include the selection of the random noise variable ($x_3$) as an important variable in many predictions, which should not be the case. The pattern seen in the explanations for 3 to 6 quantile bins may suggest local explanations, but the dependence on tuning parameters makes it unclear which set of explanations to use.

@@

The feature heatmap for the \data \ in the previous section shows an example of inconsistent LIME explanations across tuning parameter values. In this situation, the user must determine which set of explanations to trust. One way to do this is to compute assessment metrics for each set of explanations to identify the optimal tuning parameter values. We discuss three metrics for this purpose and present a visual comparison in an \emph{assessment metric plot}.

Each metric presented below is computed on  LIME explanations for a set of predictions obtained using the same tuning parameter values. Here we provide a high level description of the metrics. Notation and formulas for these metrics are included in Appendix \ref{metric-details}.

@@

\includegraphics[width=3.25in]{./figure-static/signatures.png}
\caption{Example bullet signatures. The bullet signatures correspond to the same land and therefore have very similar patterns. The \citet{hare:2016} random forest is fit using various features that measure the similarity between two such signatures.}
\label{fig:signatures}
\end{figure}

@@

We apply the \kge{trained} random forest \kge{model} from \citet{hare:2016} to another set of bullets from the Hamby study with \Sexpr{dim(bullet_test)[1]} rows of land comparisons. We first consider a global visualization of the relationship between the random forest predictions and the model features with a parallel coordinate plot of the training data (top facet) and testing data (bottom facets) predictions (\autoref{fig:bullet-plot-rfpcp}). The majority of same source observations with random forest probabilities close to 1 have a clear pattern of corresponding feature values. The known same source observations where the random forest is wrong (does not return probabilities close to 1) have feature values that reflect those of observations where same source is known to be false.

LIME is applied to all test set observations using different tuning parameter values: 12 sampling methods (2-6 equally spaced bins, 2-6 quantile bins, kernel density estimation, and normal approximation) and 3 Gower exponents (0.5, 1, and 10). Thus, a total of $12\times 3=36$ different applications of LIME are performed. We specify that each LIME explanation return 3 features and feature selection is performed by selecting the features with the highest weights in a ridge regression model (default option).

@@

First, the density simulation methods produce the same explanations for almost all cases and LIME tuning parameter values suggesting the LIME explanations are global and not local. Second, within a bin based simulation method, the features selected by LIME for an observation often vary by the number of bins but do not appear to vary by the Gower power. With the equal bins, there are vertical stripes that suggest a dependence of the LIME explanations on the number of bins. The vertical stripes are not as apparent with the quantile bins. Lastly, there are clear differences between the LIME explanations produced by the bin based simulation methods for the matches and non-matches. This suggests that the features the random forest uses to classify a match or non-match are different.

@@

@
\caption{Feature heatmap of  36 LIME applications to the bullet comparison data test set. In addition to faceting the results by simulation method and LIME feature selection order, facets for the Gower power and whether the observation is a match or non-match are included. The vertical stripes of features selected indicate a dependence between the LIME explanations and tuning parameter values.}
\label{fig:bullet-feature-heatmap}

@@

To try to identify a set of LIME tuning parameter values with the most trustworthy set of explanations, an assessment metric plot is considered (\autoref{fig:bullet-assess-metric}). All density simulation methods performed well based on the three metrics. This is an interesting result, since \autoref{fig:bullet-feature-heatmap} shows that the density methods results in global explanations. The metric values for the bin based methods do not agree across metrics. For example, 2 and 3 quantile bins performed well according to the average $R^2$ but poorly according to MSEE. As a result, with the bin based methods, it is difficult to know which method to recommend. The applications using a power of 0.5 perform the best or as well as the other powers across all simulation methods suggesting that the power that leads to a more global explanation is preferred by LIME. Perhaps we can say that the kernel density method \kge{or} 3 equal bins \sout{, or 3 quantile bins} are \kge{the best} \sout{some of the better} performing methods, but unfortunately, \autoref{fig:bullet-assess-metric} leaves us without a clear indication of which set of explanations to use to explain the random forest.

\kge{To provide a more detailed view of explanations obtained using different tuning parameter values,} \sout{To better understand the explanations provided by these three methods,} we take a closer look at explanations \sout{from these methods} for two observations of interest, which will be referred to as case M (a known match) and case NM (a known non-match)\kge{, using explanation scatterplots}. Figure \ref{fig:bullet-eois} includes the visual representation plots of explanations (from \emph{lime}) and explanation scatterplots for cases M and NM. Plots for 3 equal bins and 3 quantile bins are included \kge{to depict tuning parameter values with good (3 equal bins) and mediocre (3 quantile bins) performance based on the assessment metric plot}. Plots for the kernel density simulation method are included in Appendix~\ref{exp-scatter_plus} (Figure \ref{fig:bullet-expl-scat-density}). \sout{The scatterplots provide several insights into the LIME explanations. Here, we focus on the explanation for case NM with 3 quantile bins, but similar statements can be made about the other plots.}
